Sparse Matrix Vector Multiplication with Multiple vectors and Multiple Matrices on the MIC Architecture

### 2013-10-16 ieee_paper.tex
intro

\section{Introduction}
The multiplication of sparse matrix by a dense vector (SpMV) is an
important kernel in many fields from scientific computing~\cite{} \todo{Do you have a good citation for that?}
to graph analysis (recommender systems~\cite{Brin98}, graph
drawing~\cite{Koren05}). So improving the performance of SpMV has
captured the interest of many researcher; including the development of
various implementations for multi-core
CPUs~\cite{Buluc2009SPAA,Williams07} and GPUs~\cite{Bell08,
  conf/ipps/KreutzerHWFBB12,
  journals/concurrency/VazquezFG11,kumar2012accelerating}. The main
challenge that is faced to obtain good performance for SpMV is that
the operation are conducted using memory location that are irregular
and often unpredictible. That make that the kernels are mostly memory
bound and there is a significant instruction overhead per flop.

Common improvement techniques such as register blocking, bandwidth
reduction (matrix reordering~\cite{Cuthill69}), partitioning to fit in
cache or TLB~\cite{Nishtala07,Temam:1992:CBS:147877.148091,conf/ppsc/Toledo97},
unrolling~\cite{Mellor-Crummey04} have impacts which are very
dependent on the matrix and overall do not lead to dramatic
improvement. (The state-of-the-art techniques in OSKI~\cite{Vuduc05}
provide some useful yet limited improvements). Register blocking~\cite{conf/ppsc/Toledo97} does
not apply well to most the matrix at hand (despite it can be applied
with virtually no overhead in all cases thanks to compressed
representation~\cite{Buluc11}). there is about 8 bytes of the matrix to move
in per non zero (assuming single precision); each nonzero requires two
floating point operations leading to a flop-to-byte ratio of at most
$\frac{1}{4}$. This limits the obtained performance to at most a
quarter of the bandwidth of the architecture wasting a lot of
potentially useful cycles. The commonly used techniques are mostly
designed to reach that bound rather than overcome it.

Fortunately that fate is not inevitable. One solution would be to pair
multiple component of an application to schedule a more instruction
intensive kernel simultaneously with SpMV, relying on some hardware
threading capabilities, such as HyperThreading, to reduce the cycle
wastage. However most ot the applications that use SpMV do not
typically have an instruction intensive kernel to run simultaneously.

An other solution, and the one we pursue in this paper, is to compute
multiple SpMV at once on matrices that have the same sparsity
patterns. Obviously not all the applications have such a property. But
important classes of applications such as graph
recommendation~\cite{Kucuktunc13-SNAM}, eigensolving~\cite{LOBPCG} and the computation of
derivative of Radial Basis Functions(RBF)~\cite{} \todo{a good citation?} can use multiple
SpMVs simultaneously. In this paper in particular, we investigate the
case of the derivative of RBFs where four derivatives of four
different function is expressed as the multiplication of four vectors
by four matrices with identical sparsity patterns leading to the
simultaneous execution of 16 SpMVs at a time. Using multiple vectors
at a time has been investigated before to densify the
computations~\cite{Im01} but this is the first time that the
densification comes from adding both vectors and matrices.

@@

In Section~\ref{sec:rbf} we present the computation of RBFs and how it
can be expressed 16 multiplication of 4 vectors by 4 sparse matrices
with a common sparsity pattern. Section~\ref{sec:model} presents an
estimation of the instruction intensity of various form of the
computations and we show that a 5- to 6- fold improvement can be
expected when computing the 16 multiplications simultaneously and
reach a total of about 200 Gflop/s. This performance represents
approximatively 10\% of the available flop/s of a Knight Corner
coprocessor. Therefore, it is necessary to have implementations that
perform the computation in as little amount of instructions as
possible. We describe in Section~\ref{sec:impl} the details of the MIC
architecture and how to use specialized load, store, swizzle and
permutation instruction to efficiently bring the data in the vector
registers to be processed. Section~\ref{sec:expe} gives some
experimental result about the amount of bandwidth that can be achieved
depending on how the spmv kernel is written and the actual performance
of the various kernel on multiple classes of matrices some generated
for analysis purpose and some extracted from and RBF application. A
performance of xxx GFlop/s is achieved on real scenario. Concluding
remarks and perspectives are provided in Section~\ref{sec:ccl}.

>>>>

\section{Introduction}
The multiplication of sparse matrix by a dense vector (SpMV) is an
important kernel in many fields from scientific computing~\cite{} \todo{Do you have a good citation for that?}
to graph analysis (recommender systems~\cite{Brin98}, graph
drawing~\cite{Koren05}). So improving the performance of SpMV has
captured the interest of many researcher; including the development of
various implementations for multi-core
CPUs~\cite{Buluc2009SPAA,Williams07} and GPUs~\cite{Bell08,
  conf/ipps/KreutzerHWFBB12,
  journals/concurrency/VazquezFG11,kumar2012accelerating}. The main
challenge that is faced to obtain good performance for SpMV is that
the operation are conducted using memory location that are irregular
and often unpredictible. That makes the kernels mostly memory
bound and there is a significant instruction overhead per flop.

Common improvement techniques such as register blocking, bandwidth
reduction (matrix reordering~\cite{Cuthill69}), partitioning to fit in
cache or TLB~\cite{Nishtala07,Temam:1992:CBS:147877.148091,conf/ppsc/Toledo97},
unrolling~\cite{Mellor-Crummey04} have impacts which are very
dependent on the matrix and overall do not lead to dramatic
improvement. (The state-of-the-art techniques in OSKI~\cite{Vuduc05}
provide some useful yet limited improvements). Register blocking~\cite{conf/ppsc/Toledo97} does
not apply well to most the matrix at hand (despite it can be applied
with virtually no overhead in all cases thanks to compressed
representation~\cite{Buluc11}). There is about 8 bytes of the matrix to move
in per non zero (assuming single precision); each nonzero requires two
floating point operations leading to a flop-to-byte ratio of at most
$\frac{1}{4}$. This limits the obtained performance to at most a
quarter of the bandwidth of the architecture wasting a lot of
potentially useful cycles. The commonly used techniques are mostly
designed to reach that bound rather than overcome it.

Fortunately that fate is not inevitable. One solution would be to
schedule a more instruction intensive kernel simultaneously with the
execution of SpMV, relying on some hardware threading capabilities,
such as HyperThreading, to reduce the cycle wastage. However, most ot
the applications that use SpMV do not typically have an instruction
intensive kernel to run simultaneously.

An other solution, and the one we pursue in this paper, is to compute
multiple SpMV at once on matrices that have the same sparsity
patterns. Obviously not all the applications have such a property. But
important classes of applications such as graph
recommendation~\cite{Kucuktunc13-SNAM}, eigensolving~\cite{LOBPCG} and the computation of
derivative of Radial Basis Functions(RBF)~\cite{} \todo{a good citation?} can use multiple
SpMVs simultaneously. In this paper in particular, we investigate the
case of the derivative of RBFs where four derivatives of four
different functions is expressed as the multiplication of four vectors
by four matrices with identical sparsity patterns leading to the
simultaneous execution of 16 SpMVs at a time. Using multiple vectors
at a time has been investigated before to densify the
computations~\cite{Im01} but this is the first time that the
densification comes from adding both vectors and matrices.

@@

In Section~\ref{sec:rbf} we present the computation of RBFs and how it
can be expressed 16 multiplication of 4 vectors by 4 sparse matrices
with a common sparsity pattern. Section~\ref{sec:model} presents an
estimation of the instruction intensity of various form of the
computations and we show that a 6- fold improvement can be
expected when computing the 16 multiplications simultaneously and
reach a total of about 200 Gflop/s. This performance represents
approximatively 10\% of the available flop/s of a Knight Corner
coprocessor. Therefore, it is necessary to have implementations that
perform the computation in as little amount of instructions as
possible. We describe in Section~\ref{sec:impl} the details of the MIC
architecture and how to use specialized load, store, swizzle and
permutation instruction to efficiently bring the data in the vector
registers to be processed. Section~\ref{sec:expe} gives some
experimental result about the amount of bandwidth that can be achieved
depending on how the SpMV kernel is written and the actual performance
of the various kernel on multiple classes of matrices some generated
for analysis purpose and some extracted from and RBF application. A
performance of 135 Gflop/s is achieved on real scenario. Concluding
remarks and perspectives are provided in Section~\ref{sec:ccl}.
